Constructing a universe for the setoid model

### 2020-06-30 paper.tex
More paper writing

\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\mType}{\mathbf{Type}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\mProp}{\mathbf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Sub}{\mathrm{Sub}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}
\newcommand{\Uty}{\mathds{U}}
\newcommand{\EL}{\mathds{E}l}

@@

\section{$MLTT^{Prop}$}\label{metatheory}

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. We employ Agda
notation for metatheoretic terms throughout the paper.

One aspect of the metatheory that we pay attention to is it's minimality, as we
want to keep it as simple as possible. In particular, we want to avoid having to
assume strong definition schemas that go beyond inductive families.

$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via $\textsf{Lift} : \mProp \to \mType$.

%% The universe of propositions is
%% inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
%% \mProp$.

@@

In addition to the above type formers in $\mType$, we will need the
propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
be defined from their $\mType$ counterparts via truncation. That is, given $P :
\mProp$ and $Q : P \to \mProp$:

@@

Finally, we will assume the presence of an identity type in the style of
Martin-L\"of's inductive identity type. The main difference is that our identity
type is a $\mProp$-valued relation. Naturally, we also have a \textsf{transp}
combinator representing the corresponding elimination principle.

@@

Note that \textsf{transp} provides a strong elimination principle allowing to
eliminate a strict proposition (the identity type) into arbitrary types. Indeed,
the reason why we must assume this identity type as primitive is because, even
though we could have defined \textsf{Id} from $W_p$, \textsf{transp} would not
be definable from $W_p$ elimination. \todo{what is the justification for this?}

@@

A universe of strict propositions has been recently added to the Agda proof
assistant \cite{agda-prop}, thus making $MLTT^{Prop}$ a subset of Agda. In fact,
most of the developments presented here have been formalized and proof-checked
using Agda extended with strict propositions.
%
For convenience, we slightly deviate from $MLTT^{Prop}$ both in the paper and in
the formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $MLTT^{Prop}$, although we
haven't fully checked all the necessary details.

@@

By \emph{setoid model} we intend a class of models of Type Theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional Type Theory by M. Hofmann in
\cite{hofmann}, in order to give a semantics to extensionality
principles such as function and propositional extensionality. Here we consider
a similar model construction due to Altenkirch \cite{setoid99}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory which includes a strict universe of propositions.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates function extensionality,
a univalent universe of propositions, and quotient types. Therefore, it provides
a way to bootstrap and ``explain'' extensionality, since the model construction
effectively gives an implementation of various extensionality principles it in
terms of a small, entirely intensional theory.

@@

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers \cite{internal-tt}. The core structure of a CwF can be given as the
following signature:

@@

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & ... \\
    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

This definition of types in the setoid model is different from the one in
\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
difference here is in the use of heterogeneous equality in the definition of
types.

@@

\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : ...
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]

As defined, we can show that the setoid model validates the usual basic type
formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
of strict propositions satisfying propositional extensionality
\cite{setoid99}. Note that we don't need neither identity types nor inductive
types (W-types) in the metatheory to achieve this. However, W-types are needed
when constructing the setoid universe in the next sections.

@@

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.
\footnote{%
Semantically, this translation corresponds to a model construction, in
particular a functor from the category of models of the target theory to the
category of models of what will be Setoid Type Theory.  Since the setoid
translation is structural in the context component, we can work with models in
the style of categories with families rather than contextual categories.
}

A setoid translation is used in \cite{mpc19} to justify Setoid Type Theory
(SeTT), an extension of $MLTT^{Prop}$ (without W-types) with equality types for
contexts and dependent types that reflect the setoid equality of the model in
the syntax.

We recall the rules of SeTT that extend $MLTT^{Prop}$ below, but with a
variation. Whereas the equality types in \cite{mpc19} are stated as elements of
the universe of propositions internal to SeTT itself, we here state the context
equalities as elements of the external, metatheoretic universe
$\mProp$. Equality on dependent types is defined as before.

@@

Notably, equality constructors in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%
\begin{align*}
  & (\Pi\ A\ B)^\sim\ p\ f_0\ f_1 = ... \\
  & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
  & \Prop^\sim\ p\ P\ Q = ...
\end{align*}

@@

Note that contractibility of singletons holds in SeTT by UIP and the computation
rules of equality on $\Sigma$ types. Hence, Martin-L\"of's J eliminator can also
be derived \todo{add citation}.

@@

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section TODO and shown originally in
\cite{setoid99} is that it can be carried out in a very small type theoretic
metatheory, thus providing a way to reduce extensionality to a small intensional
core.
%
We would like to stay faithful to this ideal when constructing this setoid
universe.

A known and established method for defining universes in Type Theory relies on
induction-recursion (IR), a definition schema developed by Dybjer
\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
throughout the literature, from the already mentioned type theoretic universes,
including the original formulation \`{a} la Tarski by Martin-L\"of
\cite{libretto}, to metamathematical tools like computability predicates.

Although universe constructions in Type Theory---including our own setoid
universe---are naturally presented as inductive-recursive definitions, they
may not necessarily require a metatheory that supports induction-recursion. In
fact, it is possible to reduce some instances of induction-recursion to plain
induction (more specifically, inductive families), including some universe
definitions. We recall this reduction in Section~\ref{ir-universes}.

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such universe eliminator.
%
The question arises because we only define a weak eliminator for our final
inductive-inductive encoding of the setoid universe, for reasons that are
explained in Section~\ref{inductive-universe}. Although not currently needed, a
stronger eliminator might be necessary to justify universe induction. This
problem should not arise in the other encodings of the setoid universe (as given
in Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

\subsection{Inductive-recursive universes in Type Theory}\label{ir-universes}

@@

Induction-recursion is arguably a nicer and more natural way to define
internal universes in Type Theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\Set$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflecs the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible sets. Hence
this kind of construction requires a metatheory with at least one universe.

\subsection{A first definition of the setoid universe}\label{ir-setoid-universe}

@@

[todo... explanation]
%
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}

@@

We can show that $\Uty$ is closed under $\Pi$ types and \textsf{Bool}, including
$\beta$ and $\eta$ laws. Moreover, all the necessary equations hold
definitionally in the metatheory.

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to understood notions of
induction-recursion like in \cite{dybjer-setzer}. Nevertheless, we would like to
avoid extending the metatheory with any forms of induction-recursion, as per our
goal to keep the metatheory as small and essential as possible.

As a consequence, in the next section we transform our current
inductive-recursive definition to one that doesn't use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction trick is applied to an
inductive-recursive type of this kind.

\subsection{Inductive-inductive universe}\label{ii-universe}

We will follows the same method outlined in Section~\ref{ir-universes}. In
addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
$\ad{in-U∼}$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
\end{align*}

Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
mutually defined, they form an instance of \emph{induction-induction}, a schema
that allows the definition of a type mutually with other types that contain the
first one in their signature \cite{ii}. The main example of induction-induction
can be found in intrinsic definitions of Type Theory in Type Theory
\cite{tt-in-tt}.

@@

\subsection{Inductive universe}

@@

The reduction described in the previous section works generically for an
arbitrary finitary inductive-inductive type, thus giving a systematic way to
reduce finitary inductive-inductive definitions to inductive families. However,
it is not clear whether this method extends to infinitary induction-induction,
of which the setoid universe defined in Section~\ref{ii-universe} is an
instance.
%
Of course, the absence of a general reduction method does not mean that we can't
reduce particular concrete instances of infinitary induction-induction, which is
exactly what we hope for our universe construction.

The obvious challenge in successfully completing this reduction is to avoid the
need for extensionality in the metatheory. In fact, consider the simple
infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
soon as we try to define the general eliminator, since it requires function
extensionality. \todo{Is this true?}

Luckily, in our case we can take advantage of our specific circumstances. First
of all, we can define the well-typing predicates as $\mProp$-valued families,
rather than in $\mType$:

\begin{align*}
  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
\end{align*}
\todo{check that these signatures are correct}

>>>>

\newcommand{\mlttp}{\textsf{MLTT}^{\mProp}}
\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\mType}{\mathbf{Type}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\mProp}{\mathbf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Sub}{\mathrm{Sub}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}
\newcommand{\Uty}{\mathds{U}}
\newcommand{\EL}{\mathds{E}l}

@@

\section{$\mlttp$}\label{metatheory}

This section describes $\mlttp$, our ambient metatheory. We employ Agda notation
to write down $\mlttp$ terms throughout the paper.

One of the main appeals of Altenkirch's setoid model is that it can justify
several useful extensionality principles while being defined in a small
intensional metatheory.
%
We tried to stay true to this idea when figuring out the necessary
metatheoretical tools for the universe construction in this paper. In
particular, we wanted to avoid having to assume strong definition schemas that
go beyond inductive families.

$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:

@@

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We also include a transport combinator
\textsf{transp}.

@@

The \textsf{transp} combinator provides a strong elimination principle allowing
to eliminate a strict proposition (the identity type) into arbitrary types.
%
The reason to have an identity type in $\mProp$ with such a strong transport
rule are made clear in Section~\ref{inductive-universe}.
%
Note that this identity type must necessarily be assumed as primitive; in fact,
even though we could have defined \textsf{Id} from $W_p$, \textsf{transp} is not
definable from $W_p$ elimination.

@@

A universe of strict propositions has been recently added to the Agda proof
assistant \cite{agda-prop}, making \emph{most} of $\mlttp$ a subset of Agda,
with the exception of the proof-irrelevant identity type. Most of the universe
constructions presented here have been formalized and proof-checked using Agda,
with the proof-irrelevant identity type and the strong transport rule added via
postulates and rewriting.

For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $\mlttp$, although we haven't
fully checked all the necessary details.

@@

By \emph{setoid model} we intend a class of models of type theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional type theory by M. Hofmann in
\cite{hofmann}, in order to give a semantics to extensionality principles such
as function and propositional extensionality.

Here we consider a similar model construction due to Altenkirch
\cite{setoid99}. The peculiarity of this model is that it is presented in a type
theoretic and intensional metatheory which includes a strict universe of
propositions.

The setoid model thus defined validates function extensionality, a universe of
propositions with propositional extensionality, and quotient types. Therefore,
it provides a way to bootstrap and ``explain'' extensionality, since the model
construction effectively gives an implementation of various extensionality
principles it in terms of a small, completely intensional theory.

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:

@@

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    & \textsf{trans*} : \forall\{\gamma_0\,\gamma_1\,\gamma_2\,a_0\,a_1\,a_2\}\{p_0 : \GG^\sim\ \gamma_0\ \gamma_1\}\{p_1 : \GG^\sim\ \gamma_1\ \gamma_2\}\\
    & \ \qquad \to A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

This definition of types in the setoid model is different from the one in
\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
difference here is in the use of a heterogeneous equivalence relation $A^\sim$
in the definition of types.

@@

\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need neither identity types nor inductive types (W-types) to do this
much.

@@

The setoid model presented in the previous section is \emph{strict}, that is,
every equation of a CwF holds by definition in the semantics. One advantage of
strict models is that they can be turned into \emph{syntactic translations}, in
which syntactic objects of the source theory are interpreted as their
counterparts in another \emph{target} theory. In the case of the setoid model,
this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.
\footnote{%
Semantically, this translation corresponds to a model construction, in
particular a functor from the category of models of the target theory to the
category of models of what will be Setoid Type Theory.  Since the setoid
translation is structural in the context component, we can work with models in
the style of categories with families rather than contextual categories.
}

A setoid translation is used in \cite{mpc19} to justify Setoid Type Theory
(SeTT), an extension of Martin-L\"of type theory (+ $\mProp$) with equality
types for contexts and dependent types that reflect the setoid equality of the
model.

We recall the rules of SeTT that extend regular MLTT below, but with a
variation: whereas the equality types in \cite{mpc19} are stated as elements of
SeTT's internal universe of propositions, here we state the context equalities
as elements of the external, metatheoretic universe $\mProp$. Equality on
dependent types is defined as before in \cite{mpc19}.

@@

Notably, equality types in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%% %
%% \begin{align*}
%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
%%   & \Prop^\sim\ p\ P\ Q = ...
%% \end{align*}

@@

We can also derive Martin-L\"of's J eliminator for this homogeneous identity
type.

@@

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section~\ref{setoid-model} and shown originally
in \cite{setoid99} is that it can be carried out in a very small type theoretic
metatheory, thus providing a way to reduce extensionality to a small intensional
core.
%
We would like to stay faithful to this ideal when constructing this setoid
universe.

A known and established method for defining universes in type theory relies on
induction-recursion (IR), a definition schema developed by Dybjer
\cite{ir-dybjer, dybjer-setzer}. Inductive-recursive definitions can be found
throughout the literature, from the already mentioned type theoretic universes,
including the original formulation \`{a} la Tarski by Martin-L\"of
\cite{libretto}, to metamathematical tools like computability predicates.

Although universe constructions in type theory---including our own setoid
universe---are naturally presented as inductive-recursive definitions, they may
not necessarily require a metatheory with induction-recursion. In fact, it is
possible to reduce some instances of induction-recursion to plain induction
(more specifically, inductive families), including some universe definitions. We
recall this reduction in Section~\ref{ir-universes}.

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

\subsection{Inductive-recursive universes}\label{ir-universes}

@@

Induction-recursion is arguably a nicer and more natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}

$\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \pi_1
\end{align*}

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible types in
$\mType$. Hence this kind of construction requires a metatheory with at least
one universe.

\subsection{Inductive-recursive setoid universe}\label{ir-setoid-universe}

@@

We are now ready to give a full definition of the setoid universe:
%
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}

@@

We can show that $\Uty$ is closed under $\Pi$ types and unit types, including
$\beta$ and $\eta$ laws. Moreover, all the necessary equations (including
substitution laws) hold definitionally in the metatheory.

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to well-understood notions
of induction-recursion like the one described in \cite{dybjer-setzer}. In any
case, we would like to avoid extending the metatheory with any forms of
induction-recursion, given our goal to keep the metatheory as small and
essential as possible.

In the next section we transform our current inductive-recursive definition to
one that doesn't use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction method is applied to an
inductive-recursive type of this kind.

\subsection{Inductive-inductive setoid universe}\label{ii-universe}

We will follow the same method outlined in Section~\ref{ir-universes}. In
addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
$\ad{in-U∼}$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \mType\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \mProp) \to \mType_1
\end{align*}

Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
mutually defined, they form an instance of \emph{induction-induction}, a schema
that allows the definition of a type mutually with other types that contain the
first one in their signature \cite{ii}.\footnote{The main example of
  induction-induction is the intrinsic definition of a dependent type theory in
  type theory \cite{tt-in-tt}.}

@@

\subsection{Inductive setoid universe}

This section outlines a way to encode the inductive-inductive universe of
setoids from the previous section without assuming arbitrary inductive-inductive
definitions in the metatheory.

Before turning our attention to the setoid universe, we recall the known,
systematic method to reduce finitary inductive-inductive types to inductive
families.

@@

The reduction described in the previous section works generically for an
arbitrary finitary inductive-inductive type, thus giving a systematic way to
reduce finitary inductive-inductive definitions to inductive families. However,
it is not clear whether this method extends to \emph{infinitary}
induction-induction, of which the setoid universe defined in
Section~\ref{ii-universe} is an instance.
%
Of course, the absence of a general reduction method does not mean that we can't
reduce particular concrete instances of infinitary induction-induction, which is
exactly what we hope for our universe construction.

The obvious challenge in successfully completing this reduction is to avoid the
need for extensionality in the metatheory. In fact, consider the simple
infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
soon as we try to define the eliminator. One issue is that the definition of the
eliminator relies of a proof that the well-typing predicates $\ad{in-U}_1,
\ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
%
Without further assumptions this proof can only be done by induction, and
requires function extensionality since these predicates include higher-order
constructors.

One way to get around this is to define the well-typing predicates as
$\mProp$-valued families, rather than in $\mType$:

\begin{align*}
  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
\end{align*}
